### 2009-05-26 paper.tex
edits

Due to this possibility of unobserved ancestral sequence,
the standard CYK and Inside recursions fail on the ensemble model:
that is, they contain cyclic self-references which cannot be resolved.
The intuitive reason for this is that a (potentially-unbounded) number of residues in the ancestor can be
absent in all extant sequences.
This manifests as cycles of $\Snull$ states within the multiple-sequence SCFG,
states which emit to the ancestral sequence but not to any extant sequence.
The presence of bifurcations with one or more possibly-empty child branches,
corresponding to entire substructures in the ancestral sequence that
are absent in any extant sequence, further complicates inference.
While there exists a straightforward fix for the CYK algorithm, the Inside algorithm requires more work.

Consider the CYK algorithm first.
In order to correctly perform maximum-likelihood inference, we only
require that the maximum-likelihood parse tree corresponds to the parse
returned by the CYK algorithm; the entire
probability distribution over parses need not be preserved.
Because a state path which includes a cycle of $\Snull$ states
necessarily has lower probability than an equivalent one without, we
can simply ignore such $\Snull$ cycles during CYK.
Bifurcations with empty child branches cannot be similarly ignored,
but this problem can be addressed by creating equivalent effective
direct transitions between the parents of these bifurcation states and
their non-empty children.  Once these transformations have been performed,
maximum-likelihood inference can be performed using a standard
three-sequence CYK algorithm on the ensemble model.
(Similarly, the null cycle elimination discussed in computational
linguistics requires only that the CYK parse be preserved \cite{GecsegSteinby97}.)

The transformations just described to make CYK work properly are insufficient for the
Inside and Outside algorithms.
For CYK, null cycles can be ignored, since they should never appear in the maximum likelihood parse.
For Inside and Outside, null cycles cannot be ignored, since they will contribute to the likelihood sum over parse trees.

There are two possible solutions to this difficulty: we can either
create new versions of the standard inference algorithms for SCFGs
which can cope with SCFGs with null cycles or we can modify the
structure of the SCFGs generated by our model-construction algorithm
such that they are free of null cycles.  The first approach of
creating new algorithms is simple in principle---we can simply
iteratively re-fill entries of the DP matrix, corresponding to summing
the probabilities for unobserved null cycles, until the approximation
is sufficiently good---but it is difficult to implement efficiently
and necessarily has worse time complexity than the standard inference
algorithms due to the need to iterate over parts of the (large) DP
matrix.  The second approach, of modifying the structure of the SCFGs
themselves, has a closed-form solution for the case of HMMs, but to our
knowledge, efficient exact null-cycle elimination for SCFGs is an open
problem (complicated by the presence of null cycles
involving bifurcations).

We therefore developed a method to remove null cycles from
the ensemble model, described below, in order to use the standard inference algorithms
for SCFGs without incurring the further computational complexity
required by modifying the DP algorithms themselves.

@@

As noted above, the ensemble grammar contains many rules that can be applied redundantly,
together or in isolation, to generate subtrees of the parse tree that do
not generate any terminals, and could be replaced with simpler rules.
In fact, this is a general property of SCFGs with null states and/or bifurcations.
In this subsection, we describe how to transform the SCFG so as to eliminate such redundant rules.
We begin by identifying two types of redundant parse-structures: {\bf empty bifurcations} and {\bf empty paths}.

@@

Here, we show how null cycles can be removed systematically from SCFGs by two steps:
\begin{description}
\item[(i)] separating bifurcations
into those which have one or more empty children (and can therefore be
represented using transition or termination rules) and those that have
two nonempty children;
\item[(ii)] replacing all empty paths through null states with effective direct transitions between non-null states,
obtaining sum-over-paths probabilities by inverting the grammar's transition matrix.
\end{description}

@@

Let $G = ( \Omega, \gnons, \grules, P )$ be a stochastic context-free grammar (SCFG)
consisting of
a set of terminal symbols $\Omega$,
a set of nonterminal symbols (a.k.a. ``states'') $\gnons$,
a set $\grules$ of {\em production rules} $L \to R$ (where $L \in \gnons$ and $R \in \grhs$)
and a non-negative weight function on the rules, $P:(\gnons \times \grhs) \to [0,\infty)$.

(The rule set $\grules$ can be considered redundant to the specification of the grammar, in that $\rho \notin \grules \Leftrightarrow P(\rho)=0$.
The interpretation is that $P$ is a probabilistic weight function and $\grules$ is informally the set of rules with non-zero weight.)

@@

Let $\gtrees$ be the set of all parse trees for $G$.
Suppose that parse tree $T \in \gtrees$ makes $n_T(\rho)$ uses of rule $\rho$;
then define the parse tree weight $P(T) = \prod_{\rho \in \grules} \left[ P(\rho) \right]^{n_T(\rho)}$.

@@

Let $S \in \Omega^\ast$ denote a terminal sequence.
Let $\mathrm{seq}:\gtrees \to \Omega^\ast$ be the function mapping a parse tree to its terminal sequence.
and let $\mathrm{root}:\gtrees \to \gnons$ be the function returning the root nonterminal of a parse tree.
The {\em inside probability of $S$ rooted at $A$} is $P(S|A)\ = \sum_{T: \mathrm{seq}(T) = S,\mathrm{root}(T) = A} P(T)$.
Of particular relevance to null state elimination are the probabilities $P(\epsilon|A)$,
where $\epsilon$ is the empty sequence.
These are the probabilities that a nonterminal will expire without generating any sequence.

@@

Furthermore, we say that the grammar $G$ has {\em RNA normal-form states} if
each nonterminal (state) $A \in \gnons$ takes one of the following forms:

@@

The idea is that $N'$ is a null state that is equivalent to null state $N$ for non-empty sequence.
That is, for any terminal sequence $S$,

@@

Guided by our experience with the \emph{nanos} 3' TCE and tRNA, where
the reconstruction problem was made easy by the presence of a close
outgroup, we conducted a simulation study to investigate the
dependence of reconstruction accuracy on outgroup branch length.
We simulated the evolution of RNAs under the TKFST model along
3-taxon phylogenies (with one internal node), where we kept the branch lengths of two
sibling species constant and varied the branch length of the
outgroup.  We chose the parameters of the TKFST model to
generate ``interesting'' alignments ($\sim 0.3$ substructure indels per
alignment) and considered only alignments with at least two ancestral
stems, sequences of length $\in [30, 70]$ bp, loops of length $\in [3,
10]$ bp and stems of lengths $\in [1, 20]$ bp.
We then used \indiegram\ to reconstruct maximum likelihood reconstructions of the ancestral structure and the four-way multiple alignment (ancestor + three descendants).

The results of our simulations, illustrated in \figref{simulations},
demonstrate that the accuracy of both structural
reconstruction and alignment accuracy depend on the outgroup
branch length.  However, while the difficulty of alignment inference
increases in proportion to the outgroup branch length, becoming
near-impossible for very long outgroup branch lengths, the same need
not be true for structural reconstruction accuracy.  As the branch
length of the outgroup increases, the outgroup's structure becomes increasingly decoupled from
that of its ancestor, allowing the algorithm to ``ignore'' the
information which it contains.
Similarly, while we can perfectly reconstruct the ancestral
structure given an outgroup branch length of 0 (the trivial case), we
cannot perfectly reconstruct the alignment.

While these results correspond to a best-case scenario in which we
have a perfectly-parametrized evolutionary model, recent results
demonstrating that inference of sequence alignment is
robust to mis-parametrized probabilistic models \cite{Lunter2007}
suggest that the same may be true for probabilistic inference of
ancestral structures.

@@

% Prior art
The work reported in this paper builds on substantial prior art in the areas of evolutionary modeling and ancestral reconstruction.
The reconstruction of ancient sequences was first proposed in 1963 by Pauling and Zuckerkandl \cite{PaulingZuckerkandl63};
current applications of this idea, mostly using substitution models, are surveyed in the book edited by Liberles \cite{Liberles2007}.
Many algorithms in phylogenetics implicitly reconstruct substitution histories,
whether by parsimony \cite{Edwards63,HendyPenny82} or likelihood \cite{Felsenstein81}.
There has been much effort to model the evolution of sequences related
by a tree with models permitting indels
\cite{ThorneEtal91,HeinEtal2000,Hein2001,HolmesBruno2001,SteelHein2001,Holmes2003,KnudsenMiyamoto2003,LunterSongMiklosHein2003,MiklosLunterHolmes2004,LunterEtAl2004,BradleyHolmes2007,SatijaEtAl2008}.
Recent work has extended these ideas to the reconstruction of indel
histories \cite{KimSinha2007,DialloEtAl2007},
particularly at the genomic scale \cite{MillerHausslerEtAl2006,PatenHolmesBirney2008}.
There is substantial prior work in computational linguistics
on the theory of transducers for sequences \cite{MohriPereiraRiley2000}
and parse trees \cite{Rounds70,Thatcher70,GecsegSteinby97,ComonEtAl2007-TreeTransducers}
(from which we take the terms ``string transducer'' and ``parse-tree transducer'').
We draw on the bioinformatics literature for SCFGs
\cite{Eddy94,Sakakibara94c,Durbin98},
especially Pair SCFGs \cite{RivasEddy99,Holmes2005,DowellEddy2006}
and phylogenetic SCFGs \cite{KnudsenHein2003}.
In particular, an early example of a pairwise conditional model $P(Y|X)$ for structure-dependent RNA evolution
may be found in \cite{KleinEddy2003}.
A conditional framework similar to ours in some respects is described
by Sakakibara et al \cite{Sakakibara2003}.
The statistical inference algorithms for multiple-sequence SCFGs are closely related to the ``protosequence algorithm'' of Sankoff \cite{Sankoff85},
a seminal work in both RNA structural alignment and ancestral reconstruction.

@@

Reconstructing structural changes of large RNAs
using the three-way sampling kernel which we have described
would require resources far in excess of what are currently available;
barring the availability of supercomputers with terabytes of memory,
such algorithms will only be feasible for short RNAs (\tabref{complexities}).
Thus, alternate strategies will have to be explored, such as the
progressive reconstruction used by the \ortheus\ program for
reconstructing ancestral genomes \cite{PatenEtAl2008}.
Another promising direction is to consider variations on the three-way sampling kernel,
such as the importance-sampling approach described for the TKF model
by \cite{RedelingsSuchard2005}.
This approach first proposes an ancestor $W$ by aligning extant
sequence $X$ to $Y$ (ignoring $Z$); then, in a second step, the proposed $W$ is
independently aligned to $Z$. The proposed three-way alignment and
reconstruction is then randomly accepted (or rejected) using a
Hastings ratio based on the three-way transducer composition. The
complexity of this kernel is the same as the pairwise case;
with suitable constraints, this is feasible for RNA
grammars on present hardware, at least for ribosomal domains (if not
yet whole subunits---although pairwise alignment of those should also
be possible soon). The approach of Redelings and Suchard therefore
merits future consideration in the context of modeling the evolution
of RNAs on a tree.

>>>>

Due to this possibility of unobserved ancestral sequence,
the standard CYK and Inside recursions fail on the ensemble model:
that is, they contain cyclic self-references which cannot be resolved.
The intuitive reason for this is that a (potentially-unbounded) number of residues in the ancestor can be
absent in all extant sequences.
This is manifested as cycles of $\Snull$ states within the multiple-sequence SCFG,
states which emit to the ancestral sequence but not to any extant sequence.
The presence of bifurcations with one or more possibly-empty child branches,
corresponding to entire substructures in the ancestral sequence that
are absent in any extant sequence, further complicates inference.
While there exists a straightforward fix for the CYK algorithm, the Inside algorithm is more complicated.

We consider the CYK algorithm first.
In order to correctly perform maximum-likelihood inference, we only
require that the maximum-likelihood parse tree corresponds to the parse
returned by the CYK algorithm; the entire
probability distribution over parses need not be preserved.
Because a state path which includes a cycle of $\Snull$ states
necessarily has lower probability than an equivalent one without, we
can simply ignore such $\Snull$ cycles during CYK.
Bifurcations with empty child branches cannot be similarly ignored,
but this problem can be addressed by creating equivalent effective
direct transitions between the parents of these bifurcation states and
their non-empty children.  Once these transformations have been performed,
maximum-likelihood inference can be performed using a standard
three-sequence CYK algorithm on the ensemble model.
(Similarly, the null cycle elimination discussed in computational
linguistics requires only that the CYK parse be preserved \cite{GecsegSteinby97}.)

The transformations just described to make CYK work properly are insufficient for the
Inside and Outside algorithms.
For CYK, parse trees containing null cycles can be ignored, since the maximum likelihood parse tree should never contain null cycles.
For Inside and Outside, such null-cycle parses cannot be ignored, since they will contribute to the likelihood sum over parse trees.

There are two possible solutions to this difficulty: we can either
create new versions of the standard inference algorithms for SCFGs
which can cope with SCFGs with null cycles or we can modify the
structure of the SCFGs generated by our model-construction algorithm
such that they are free of null cycles.  The first approach of
creating new algorithms is simple in principle---we can simply
iteratively re-fill entries of the DP matrix, corresponding to summing
the probabilities for unobserved null cycles, until the approximation
is sufficiently good---but it is difficult to implement efficiently
and necessarily has worse time complexity than the standard inference
algorithms due to the need to reiterate over (large) parts of the DP
matrix.  The second approach, of modifying the structure of the SCFGs
themselves, is well-understood for the case of HMMs, but to our
knowledge, still an open problem for SCFGs
(complicated by the presence of null cycles involving bifurcations).

We therefore developed a method to eliminate null cycles from
the ensemble model, in order to use the standard inference algorithms
for SCFGs without incurring the further computational complexity
required by modifying the DP algorithms themselves.
We will describe this elimination procedure before proceeding to the descriptions of the DP algorithms.

@@

As noted above, the ensemble grammar contains many rules that can be applied redundantly,
together or in isolation, to generate subtrees of the parse tree that do
not generate any terminals.
This generates cyclic dependencies in the standard DP recursions for inference.
In this subsection, we describe how to transform the SCFG so as to eliminate such redundant rules, yielding strictly acyclic DP recursions.
This transformation can be applied to any SCFG so as to remove null states and/or bifurcations:
the procedure is not restricted to grammars that were generated using our transducer composition algorithm.

We begin by identifying two distinct classes of redundant parse-subtree: {\bf empty bifurcations} and {\bf empty paths}.
We will eliminate each of these in turn.

@@

Null cycles can be removed systematically from SCFGs by two steps:
\begin{description}
\item[(i)] separating bifurcations
into those which have one or more empty children (and can therefore be
represented using transition or termination rules) and those that have
two nonempty children;
\item[(ii)] replacing all empty paths through null states with effective direct transitions between non-null states,
obtaining sum-over-paths probabilities by inverting the grammar's transition matrix.
\end{description}

@@

Let $G = ( \Omega, \gnons, \grules, P )$ be a stochastic context-free grammar (SCFG)
consisting of
a set of terminal symbols $\Omega$,
a set of nonterminal symbols (a.k.a. ``states'') $\gnons$,
a set $\grules$ of {\em production rules} $L \to R$ (where $L \in \gnons$ and $R \in \grhs$)
and a probability function on the rules, $P:(\gnons \times \grhs) \to [0,\infty)$.

% (The rule set $\grules$ can be considered redundant to the specification of the grammar, in that $\rho \notin \grules \Leftrightarrow P(\rho)=0$.
% The interpretation is that $P$ is a probabilistic weight function and $\grules$ is informally the set of rules with non-zero weight.)

@@

Let $\gtrees$ be the set of all parse trees for $G$.
Suppose that parse tree $T \in \gtrees$ makes $n_T(\rho)$ uses of rule $\rho$;
then define the parse tree likelihood $P(T) = \prod_{\rho \in \grules} \left[ P(\rho) \right]^{n_T(\rho)}$.

@@

Let $S \in \Omega^\ast$ denote a terminal sequence.
Let $\mathrm{seq}:\gtrees \to \Omega^\ast$ be the function mapping a parse tree to its terminal sequence.
and let $\mathrm{root}:\gtrees \to \gnons$ be the function returning the root nonterminal of a parse tree.
The {\em inside probability of $S$ rooted at $A$} is $P(S|A)\ = \sum_{T: \mathrm{seq}(T) = S,\mathrm{root}(T) = A} P(T)$.
Of particular relevance to null state elimination is the probability $P(\epsilon|A)$,
where $\epsilon$ is the empty sequence.
This is the probability that a nonterminal $A$ will expire without generating any sequence.

@@

Suppose that grammar $G$ has RNA normal-form rules.
We further say that $G$ has {\em RNA normal-form states} if
each nonterminal (state) $A \in \gnons$ takes one of the following forms:

@@

The idea is that $N'$ (in $G'$) is a null state that is equivalent to null state $N$ (in $G$) for non-empty sequence only.
That is, for any terminal sequence $S$,

@@

Guided by our experience with the \emph{nanos} 3' TCE and tRNA, where
the reconstruction problem was made easy by the presence of a close
outgroup, we conducted a simulation study to investigate the
dependence of reconstruction accuracy on outgroup branch length.
(We here use the term ``outgroup'' loosely, to denote the variable-length branch in a three-branch study,
where the other two branches are held at unit length.
Technically, the term ``outgroup'' is only correct if the outgroup's branch is longer than the other two branches.)

We simulated the evolution of RNAs under the TKFST model along
3-taxon phylogenies (with one internal node), where we kept the branch lengths of two
sibling species constant and varied the branch length of the
outgroup.  We chose the parameters of the TKFST model to
generate ``interesting'' alignments ($\sim 0.3$ substructure indels per
alignment) and considered only alignments with at least two ancestral
stems, sequences of length $\in [30, 70]$ bp, loops of length $\in [3,
10]$ bp and stems of lengths $\in [1, 20]$ bp.
We then used \indiegram\ to reconstruct maximum likelihood reconstructions of the ancestral structure and the four-way multiple alignment (ancestor + three descendants).

The results of our simulations, illustrated in \figref{simulations},
demonstrate that the accuracy of both structural
reconstruction and alignment accuracy depend on the outgroup
branch length.  However, while the difficulty of alignment inference
increases in proportion to the outgroup branch length, becoming
near-impossible for very long outgroup branch lengths, the same need
not be true for structural reconstruction accuracy.  As the branch
length of the outgroup increases, the outgroup's structure becomes increasingly decoupled from
that of its ancestor, allowing the algorithm to ``ignore'' the
information which it contains.
In the trivial case where the outgroup branch length is zero (so the ancestor is identical to the outgroup),
we can perfectly reconstruct the ancestral structure (as expected),
but we cannot perfectly reconstruct the three-way multiple alignment (since there are still two finite-length branches, on which some unreconstructable mutations inevitably occur).

These results correspond to a best-case scenario in which we
have a perfectly-parametrized evolutionary model and exactly-known descendant structures.
Recent results
demonstrating that inference of sequence alignment is
robust to mis-parametrized probabilistic models \cite{Lunter2007}
suggest that the same may be true for probabilistic inference of
ancestral structures.
The effect of imperfect knowledge of the ancestral structures has not been explored.

@@

% Prior art
The work reported in this paper builds on substantial prior art in the areas of evolutionary modeling and ancestral reconstruction.
The reconstruction of ancient sequences was first proposed in 1963 by Pauling and Zuckerkandl \cite{PaulingZuckerkandl63};
current applications of this idea, mostly using substitution models, are surveyed in the book edited by Liberles \cite{Liberles2007}.
Many algorithms in phylogenetics implicitly reconstruct substitution histories,
whether by parsimony \cite{Edwards63,HendyPenny82} or likelihood \cite{Felsenstein81}.
There has been much effort to model the evolution of sequences related
by a tree with models permitting indels
\cite{ThorneEtal91,HeinEtal2000,Hein2001,HolmesBruno2001,SteelHein2001,Holmes2003,KnudsenMiyamoto2003,LunterSongMiklosHein2003,MiklosLunterHolmes2004,LunterEtAl2004,BradleyHolmes2007,SatijaEtAl2008}.
Recent work has extended these ideas to the reconstruction of indel
histories \cite{KimSinha2007,DialloEtAl2007},
particularly at the genomic scale \cite{MillerHausslerEtAl2006,PatenHolmesBirney2008}.
There is substantial prior work in computational linguistics
on the theory of transducers for sequences \cite{MohriPereiraRiley2000}
and parse trees \cite{Rounds70,Thatcher70,GecsegSteinby97,ComonEtAl2007-TreeTransducers}
(from which we take the terms ``string transducer'' and ``parse-tree transducer'').
We draw on the bioinformatics literature for SCFGs
\cite{Eddy94,Sakakibara94c,Durbin98},
especially Pair SCFGs \cite{RivasEddy99,Holmes2005,DowellEddy2006}
and phylogenetic SCFGs \cite{KnudsenHein2003}.
In particular, an early example of a pairwise conditional model $P(Y|X)$ for structure-dependent RNA evolution
may be found in \cite{KleinEddy2003}.
A conditional framework similar to ours in some respects is described
by Sakakibara et al \cite{Sakakibara2003}.
The dynamic programming inference algorithms for multiple-sequence SCFGs are closely related to the ``protosequence algorithm'' of Sankoff \cite{Sankoff85}.
%a seminal work in both RNA structural alignment and ancestral reconstruction.

@@

Reconstructing structural changes of large RNAs
using the three-way sampling kernel which we have described
would require resources far in excess of those currently available;
barring the availability of supercomputers with terabytes of memory,
such algorithms will only be feasible for short RNAs (\tabref{complexities}).
Thus, alternate strategies will have to be explored, such as the
progressive reconstruction used by the \ortheus\ program for
reconstructing ancestral genomes \cite{PatenEtAl2008}.
Another promising direction is to consider variations on the three-way sampling kernel,
such as the importance-sampling approach described for the TKF model
by \cite{RedelingsSuchard2005}.
This approach first proposes an ancestor $W$ by aligning extant
sequence $X$ to $Y$ (ignoring $Z$); then, in a second step, the proposed $W$ is
independently aligned to $Z$. The proposed three-way alignment and
reconstruction is then randomly accepted (or rejected) using a
Hastings ratio based on the three-way transducer composition. The
complexity of this kernel is the same as the pairwise case;
with suitable constraints, this is feasible for RNA
grammars on present hardware, at least for ribosomal domains (if not
yet whole subunits---although pairwise alignment of those should also
be possible soon). The approach of Redelings and Suchard therefore
merits future consideration in the context of modeling the evolution
of RNAs on a tree.
